Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-05 tagless/tagless-final.tex
Comments on Sec 2 and 3

Because the types in the |Symantics| class reflect the object language's type
system, this representation admits only well-typed object terms for evaluation.
If we make a mistake somewhere, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is compiled away.
In both OCaml and Haskell, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
That is, the proposition below follows immediately from the soundness of
the metalanguage's type system.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}

@@

>>>>

The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete code.

Because the types in the |Symantics| class reflect the object language's type
system, this representation admits only well-typed object terms for evaluation.
\oleg{replace the above with the following?}
The comparison of |Symantics| with Figure~\ref{fig:object} clearly
shows that every typed, closed object term can be represented in the
metalanguage. Moreover,
\begin{prop}
If an object term has the object type |t|, its
representation in the metalanguage has the type 
|forall repr. Symantics repr => repr t|
\oleg{or, has the type repr t for some repr. Meta-language embeds all
  object types.}
\end{prop}
If we make a mistake somewhere, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
That is, the proposition below follows immediately from the soundness of
the metalanguage's type system.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}
In general, 
\begin{prop}
  If the definitional interpreter implementing |Symantics| 
  in the soundly typed metalanguage never gets stuck,
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  the interpreter.
\end{prop}

@@

The overarching theme of our approach -- representation of object terms in 
\S\ref{ourapproach}, future-stage code in \S\ref{S:PE-lift} and static
values in \S\ref{S:PE-solution} and CPS values in \S\ref{S:CPS} --
demonstrates that pushing the computation towards the producers
\oleg{or, constructors. Ken said: functionalization} generally
requires simpler type systems (such as mere HM). The same idea appears
in many other areas (e.g., ammortized heap analysis, TFP/TLCA paper).
\oleg{I was told that speculation and informality is permitted in
  conclusions.}
